Two years of CiTO annotations

### 2022-11-13 article.Rmd
Description of a reuse networks, and some layout tweaks

Citations are an essential aspect of research communication and have become the basis of many
metrics in the academic publishing world. Some see citation counts as a mark of scientific impact or
even quality, but in reality the reasons for citing other work are manifold.
Two years ago the *Journal of Cheminformatics* proposed a pilot for the adoption of annotating
citations with their annotations. Basically, when you cite a source like a journal
article or datasets, you also explain why specifically you cite that article. Particularly,
the agreement and disagreement and reuse of methods and data is interesting.
This article explores what happened after the launch of the pilot, gives examples of
how people used the idea, and what needs to happen next.

@@

## Wikidata and Scholia

@@

@@

## Adoption by *Journal of Cheminformatics* authors

@@

## Technological innovation

@@

## Discussion

@@

## What is next?

>>>>

Citations are an essential aspect of research communication and have become the basis of many
metrics in the academic world. Some see citation counts as a mark of scientific impact or
even quality, but in reality the reasons for citing other work are manifold which makes
the interpretation more complicated than a single citation count can reflect.
Two years ago, the *Journal of Cheminformatics* proposed the CiTO Pilot for the adoption of
a practice of annotating citations with their citation intentions.
Basically, when you cite a journal article or dataset (or any other soruce), you also explain
why specifically you cite that source. Particularly,
the agreement and disagreement and reuse of methods and data are of interest.
This article explores what happened after the launch of the pilot. We summarize how
authors in the *Journal of Cheminformatics* used the pilot, shows citation annotations
are distributed with Wikidata, visualized with Scholia, discusses adoption outside BioMedCentral,
and ponders about what needs to happen next.

@@

# Wikidata and Scholia

@@

# Adoption by *Journal of Cheminformatics* authors

@@

@@

# Technological innovation

@@

# Annotated Citation Networks

We already use citation networks in finding relevant literature, for example based on co-citation patterns.
Such analyses become stronger when we know more why articles cited. Similarly, an article that cites an
article because it uses a method in second article and that method extends a method in a third articles,
then the first article indirectly uses the method in that third article, even if that third article is
not directly cited. This is reflecting in citation habits: authors always decide whether to cite all
articles about a method, only the most recent, or only the oldest (or something else). After all, journals
frequently have rules about the maximum length of reference sections. Moreover, some methods are
so well established, we are not expected to cite that work at all.

![Citation network consisting only of citations annotated with **[cito:usesMethodIn]**.
Articles with explicit CiTO annotation are shown in red. The Wikidata SPARQL query
used to create this diagram is available from
[https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856](https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856).](images/usesMethodInNetwork.png)

The general availability of open citation allows us to recover such more complex reuse scenarios
using the citation networks. Moreover, when the citations are annotated, we can zoom in on reuse
networks. Figure 3 shows a method reuse network for articles with explicit annotation in red. The
network shows a few article that use methods in two cited articles, e.g. *Kohulan Rajan, 2021*
and *Henning Otto Brinkhaus, 2022* [@Rajan2021,@Brinkhaus2022]. Of course, if we do not limit
the network to a subset of articles (here, *Journal of Cheminformatics* articles with CiTO
annotation), the network becomes more interesting, but also much more complex. Network analysis
approaches can then be used instead of network visualization.

When we combined the reuse of methods and data, we can for journals summarize which articles
are most reused. Analyses like this becomes a simple query that can be routinely performed for
any journal. Figure 4 shows a tabular summary of articles of which methods or data is reused.
Like citation counts, this data depends totally on explicit citation data. However, these
citation counts are based on actual reuse and not also on the number of citations as authority.

![Screenshot of the "Most reused articles" section of the Scholia page for the
*Journal of Cheminformatics*. Based on CiTO annotation available in Wikdiata, the ten
most reused articles are shown.](images/mostUsedInJCheminf.png)

# Discussion

@@

# What is next?
